---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ConSciR

<!-- badges: start -->
<!-- badges: end -->

ConSciR is an R package designed to assist conservators, scientists and engineers by providing a toolkit for performing calculations and streamlining common tasks in cultural heritage conservation. ConSciR is designed for:

-   Conservators working in museums, galleries, and heritage sites
-   Conservation scientists, engineers and researchers
-   Cultural heritage professionals involved in preventive conservation
-   Students and educators in conservation and heritage science programs
-   FAIR: ConSciR is designed to be findable, accessible, interoperable and reusable

Learn more about ConSciR here:  [**ConSciR webpage**](https://bhavshah01.github.io/ConSciR/)

The ConSciR Github page: [**ConSciR Github**](https://github.com/BhavShah01/ConSciR)

If using R for the first time, read an article here: [Using R for the first time](https://bhavshah01.github.io/ConSciR/articles/ConSciR-FirstTimeR.html)

## Tools available 


-   Calculations: humidity calculations, conservator tools 
-   Graphs: TRH plots, psychometric chart
-   Summaries:  
-   Analysis: 
-   Tidying data: 



## Installation

You can install the development version of ConSciR from [GitHub](https://github.com/) with:

``` r
install.packages("pak")
pak::pak("BhavShah01/ConSciR")
```
-or- 

``` r
# install.packages("devtools")
devtools::install_github("BhavShah01/ConSciR")
```

## Examples

This is a basic example of some commons tasks:

-   Load packages

```{r packages, message=FALSE, warning=FALSE}
library(ConSciR)
library(dplyr)
library(ggplot2)
```

-   Pre-loaded dataset is availabe for testing

```{r load_dataset, message=FALSE, warning=FALSE}
# My TRH data
head(mydata)
```

-   Use functions to perform calculations, for example use the existing data to add dew point and absolute humidity. Performance metrics are also being added like lifetime multiplier, preservation index and mould calculations. 

```{r add_calcs, message=FALSE, warning=FALSE}
# Peform calculations
head(mydata) |>
  mutate(
    DewP = calcDP(Temp, RH), 
    Abs = calcAH(Temp, RH), 
    LifeTime = calcLM(Temp, RH, EA = 100), 
    PI = calcPI(Temp, RH)
    )
```

-   Combine analysis with graphs

```{r graphTRH_DewPoint, message=FALSE, warning=FALSE, fig.alt="graphTRH"}
mydata |>
  mutate(DewPoint = calcDP(Temp, RH)) |>
  graph_TRH() + 
  geom_line(aes(Date, DewPoint), col = "purple", size = 1) + 
  theme_minimal()
```

-   Use analysis functions, for example estimate mould growth

```{r mould_risk, message=FALSE, warning=FALSE, fig.alt="mould"}
mydata |>
  calcMould(Temp = "Temp", RH = "RH") |>
  ggplot() +
  geom_area(aes(Date, mould), fill = "darkgreen", size = 1) +
  labs(title = "Probabilty of mould", x = "", y = "Mould risk") + 
  theme_bw()
```

