---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ConSciR

<!-- badges: start -->
<!-- badges: end -->

ConSciR is an R package designed to assist conservators, scientists and engineers by providing a toolkit for performing calculations and streamlining common tasks in cultural heritage conservation. 

ConSciR is designed for and to be:

-   Conservators working in museums, galleries, and heritage sites
-   Conservation scientists, engineers and researchers
-   Cultural heritage professionals involved in preventive conservation
-   Students and educators in conservation and heritage science programs
-   FAIR: Findable, Accessible, Interoperable, and Reusable
-   Collaborative: allowing anyone to upload code, raise requests, report bugs, and add functionality to the package


If using R for the first time, read an article here: [Using R for the first time](https://bhavshah01.github.io/ConSciR/articles/ConSciR-FirstTimeR.html)

## Tools  

-   Humidity calculations, conservator tools, sustainability 
-   Mould growth models, damage functions
-   Graphs: TRH plots, psychometric chart
-   Tidying data: Meaco, Hanwell data
-   Interactive Shiny applications


## Installation

You can install the development version of ConSciR from [GitHub](https://github.com/) with:

``` r
install.packages("pak")
pak::pak("BhavShah01/ConSciR")
```
-or- 

``` r
# install.packages("devtools")
devtools::install_github("BhavShah01/ConSciR")
```

## Examples

This is a basic example of some commons tasks:

-   Load packages

```{r packages, message=FALSE, warning=FALSE}
library(ConSciR)
library(dplyr)
library(ggplot2)
```

-   Pre-loaded dataset is availabe for testing

```{r load_dataset, message=FALSE, warning=FALSE}
# My TRH data
head(mydata)
```

-   Use functions, for example use the existing data to add dew point and absolute humidity. Performance metrics are also being added like lifetime multiplier, preservation index and mould calculations. 

```{r add_calcs, message=FALSE, warning=FALSE}
# Peform calculations
head(mydata) |>
  mutate(
    DewP = calcDP(Temp, RH), 
    Abs = calcAH(Temp, RH), 
    LifeTime = calcLM(Temp, RH, EA = 100), 
    PI = calcPI(Temp, RH)
    )
```

-   Combine analysis with graphs

```{r graphTRH_DewPoint, message=FALSE, warning=FALSE, fig.alt="graphTRH"}
mydata |>
  mutate(DewPoint = calcDP(Temp, RH)) |>
  graph_TRH() + 
  geom_line(aes(Date, DewPoint), col = "purple") + # add dewpoint line in purple
  theme_minimal()
```

-   Conservator tools: for example estimate mould growth

```{r mould_risk, message=FALSE, warning=FALSE, fig.alt="mould"}
mydata |>
  mutate(Mould = calcMould_Zeng(Temp, RH)) |>
  ggplot() +
  geom_line(aes(Date, RH), col = "blue") +
  geom_line(aes(Date, Mould), col = "darkgreen", size = 1) +
  labs(title = "Mould Growth Rate Limits", 
       y = "Humidity (blue) and Mould Limit (green)") + 
  theme_bw()
```


-   Humidity functions: for example psychrometric chart

```{r psychart, message=FALSE, warning=FALSE, fig.alt="psych_chart"}
mydata |>
  graph_psychrometric()
```

