---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ConSciR

<!-- badges: start -->
<!-- badges: end -->

ConSciR is an R package specifically designed to assist conservators and scientists by providing a toolkit for performing calculations and streamlining common tasks in conservation. ConSciR is designed for:

-   Conservators working in museums, galleries, and heritage sites
-   Conservation scientists and researchers
-   Cultural heritage professionals involved in preventive conservation
-   Students and educators in conservation and heritage science programs

Learn more about ConSciR here:  [**ConSciR webpage**](https://bhavshah01.github.io/ConSciR/)

The ConSciR Github page: [**ConSciR Github**](https://github.com/BhavShah01/ConSciR)


## Tools available 


-   Calculations: dew point, absolute humidity, water vapour pressure, mixing ratio, enthalpy
-   Graphs: TRH plots, psychometric chart
-   Summaries: percentiles 
-   Analysis: 
-   Tidying data: 



## Installation

You can install the development version of ConSciR from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("BhavShah01/ConSciR")
```
-or- 

``` r
# install.packages("devtools")
devtools::install_github("BhavShah01/ConSciR")
```

## Examples

This is a basic example of some commons tasks:

-   Load packages

```{r load the relevant packages, message=FALSE, warning=FALSE}
library(ConSciR)
library(dplyr)
library(ggplot2)
```

-   Pre-loaded dataset is availabe for testing

```{r look at the loaded dataset, message=FALSE, warning=FALSE}
# My TRH data
head(mydata)
```

-   Use functions to perform calculations, for example use the existing data to add dew point and absolute humidity. Performance metrics are also being added like lifetime multiplier, preservation index and mould calculations. 

```{r perform calculations on the dataset, message=FALSE, warning=FALSE}
# Peform calculations
head(mydata) |>
  mutate(
    DewP = calcDP(Temp, RH), 
    Abs = calcAH(Temp, RH), 
    LifeTime = calcLM(Temp, RH, EA = 100), 
    PI = calcPI(Temp, RH)
    )
```

-   Combine analysis with graphs

```{r graph the TRH data after adding dew point, message=FALSE, warning=FALSE, fig.alt="graph the TRH data after adding dew point"}
mydata |>
  mutate(DewPoint = calcDP(Temp, RH)) |>
  graph_TRH() + 
  geom_line(aes(Date, DewPoint), col = "purple", size = 1) + 
  theme_minimal()
```

-   Use analysis functions, for example estimate mould growth

```{r use the mould risk function, message=FALSE, warning=FALSE, fig.alt="use the mould risk function"}
mydata |>
  calcMould(Temp = "Temp", RH = "RH") |>
  ggplot() +
  geom_area(aes(Date, mould), fill = "darkgreen", size = 1) +
  labs(title = "Probabilty of mould", x = "", y = "Mould risk") + 
  theme_minimal()
```

