---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ConSciR

<!-- badges: start -->
<!-- badges: end -->

ConSciR is an R package specifically designed to assist conservators and scientists by providing a toolkit for performing calculations and streamlining common tasks in conservation. ConSciR is designed for:

-   Conservators working in museums, galleries, and heritage sites
-   Conservation scientists and researchers
-   Cultural heritage professionals involved in preventive conservation
-   Students and educators in conservation and heritage science programs

Learn more about ConSciR here:  [**ConSciR webpage**](https://bhavshah01.github.io/ConSciR/)

The ConSciR Github page: [**ConSciR Github**](https://github.com/BhavShah01/ConSciR)


## Tools available 


-   Calculations: dew point, absolute humidity, water vapour pressure, mixing ratio, enthalpy
-   Graphs: TRH plots, psychometric chart
-   Summaries: percentiles 
-   Analysis: 
-   Tidying data: 



## Installation

You can install the development version of ConSciR from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("BhavShah01/ConSciR")
```
-or- 

``` r
# install.packages("devtools")
devtools::install_github("BhavShah01/ConSciR")
```

## Examples

This is a basic example of some commons tasks:

-   Load packages

```{r, message=FALSE, warning=FALSE}
library(ConSciR)
library(dplyr)
library(ggplot2)
```

-   Pre-loaded dataset is availabe for testing

```{r, message=FALSE, warning=FALSE}
# My TRH data
head(mydata)
```

-   Use functions to perform calculations, for example use the existing data to add dew point and absolute humidity. Performance metrics are also being added like lifetime multiplier, preservation index and mould calculations. 

```{r, message=FALSE, warning=FALSE}
# Peform calculations
head(mydata) |>
  mutate(
    DewPoint = calcDP(Temp, RH), 
    AbsHum = calcAH(Temp, RH), 
    LifeTime = calcLM(Temp, RH, EA = 100), 
    PreservationIndex = calcPI(Temp, RH))
```

-   Graphs 

```{r, message=FALSE, warning=FALSE}
# Produce graphs 
mydata |>
  graph_TRH()
```

-   Combine analysis with graphs

```{r, message=FALSE, warning=FALSE}
mydata |>
  mutate(DewPoint = calcDP(Temp, RH)) |>
  graph_TRH() + 
  geom_line(aes(Date, DewPoint), col = "purple", size = 1) + 
  theme_bw()
```

