---
title: "ConSciR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ConSciR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```


## Introduction

ConSciR is an R package specifically designed to assist conservators, scientists and engineers by providing a toolkit for performing calculations and streamlining common tasks in cultural heritage conservation.


## Install and load

``` r
install.packages("pak")
pak::pak("BhavShah01/ConSciR")
```
-or- 

``` r
# install.packages("devtools")
devtools::install_github("BhavShah01/ConSciR")
```

The ConSciR Github page: [ConSciR Github](https://github.com/BhavShah01/ConSciR)


## Examples 

```{r load packages, message=FALSE, warning=FALSE}
# Load packages
library(ConSciR)
library(dplyr)
library(ggplot2)
```


Add calculated values using mutate

```{r table of calculated values, message=FALSE, warning=FALSE}
# Add calculated values using mutate
head(mydata) |> 
  mutate(
    Absolute_Humidity = calcAH(Temp, RH), 
    Dew_Point = calcDP(Temp, RH), 
    Mixing_Ratio = calcMR(Temp, RH), 
    Humidity_Ratio = calcHR(Temp, RH),
    Enthalpy = calcEnthalpy(Temp, RH), 
    Saturation_Vapour_Pressure = calcPws(Temp), 
    Actual_Vapour_Pressure = calcPw(Temp, RH), 
    Air_Density = calcAD(Temp, RH),
    Temp_calc = calcTemp(RH, Dew_Point),
    RH_AH_calc = calcRH_AH(Temp, Absolute_Humidity),
    RH_DP_calc = calcRH_DP(Temp, Dew_Point)
  ) |>
  glimpse()

```


Perform calculations to visualise and explore data

```{r calculations to visualise the data, message=FALSE, warning=FALSE, fig.alt="visualisations"}

mydata |>
  # Calculate Absolute Humidity and Dew Point
  mutate(
    AbsHum = calcAH(Temp, RH),
    DewPoint = calcDP(Temp, RH)
  ) |>
  # Create base plot using graph_TRH function
  graph_TRH() +
  # Add Absolute Humidity line
  geom_line(aes(Date, AbsHum), color = "green") +
  # Add Dew Point line
  geom_line(aes(Date, DewPoint), color = "purple") +
  # Apply a theme
  theme_bw()

```

Built in psychrometric chart

```{r psychrometric chart, message=FALSE, warning=FALSE, fig.alt="psychrometric_chart_example"}

head(mydata, 100) |>
  graph_psychrometric() + 
  theme_bw()

```
