---
title: "Humidity functions"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE, 
  message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ConSciR)
library(dplyr)
library(ggplot2)
```


The relationship between the humidity functions is studied and summarised. A grid of temperature and relative humidity values, ranging from -50°C to 100°C and 0% to 100% RH respectively, was used to test the functions. Contour plots were utilized to visualize the interactions between derived variables across this temperature-humidity grid. The analysis includes validation checks for temperature and relative humidity calculations, to identify potential discrepancies. 

The humidity functions assume atmospheric pressure. 


### TRHgrid: Dataset used to visualise the functions 

A grid of temperature and relative humidity values is used to analysis the relationship between derived variables. A grid of temperature and humidity between -50 to 100°C and 0 to 100%RH with 0.25°C and 1%RH spacing. The expand.grid() function creates a complete factorial combination of all temperature and relative humidity values, where each unique temperature-humidity pair is represented..

```{r}

Temp <- seq(-50, 100, 0.25)
RH <- seq(0, 100, 1)

TRHgrid <- 
  expand.grid(Temp, RH) |> 
  tibble() |> 
  rename(Temp = Var1, RH = Var2) 

summary(TRHgrid)

head(TRHgrid, 3) 
tail(TRHgrid, 3)

```

Example of use of the functions

```{r}
TRHgrid |> 
  mutate(
    Abs = calcAH(Temp, RH), 
    DewPoint = calcDP(Temp, RH), 
    Temp_calc = calcTemp(RH, DewPoint),
    RH_AH_calc = calcRH(Temp, Abs),
    RH_DP_calc = calcRH_DP(Temp, DewPoint),
    MR = calcMR(Temp, RH), 
    Enthalpy = calcEnthalpy(Temp, RH), 
    Pws = calcPws(Temp), 
    Pw = calcPw(Temp, RH),
    LM = calcLM(Temp, RH))
```


### Temperature

The temperature calculations are explored to determine the accuracy of temperature derivation using dew point and relative humidity. Visualisation of the temperature calculation against the original temperature and relative humidity. The temperature calculation is presented as a filled contour plot of the calculated temperature using calcTemp through dew point which is calculated using calcDP. 

```{r}
head(mydata) |> dplyr::mutate(DewPoint = calcDP(Temp, RH), Temp2 = calcTemp(RH, DewPoint))

TRHgrid |>
  mutate(
    DewPoint = calcDP(Temp, RH),
    TempCalc = calcTemp(RH, DewPoint)
    ) |>
  ggplot(aes(Temp, RH, z = TempCalc)) + 
  geom_contour_filled(bins = 20, breaks = seq(-50, 100, 10)) +
  labs(title = "Calculated temperature vs actual", 
       subtitle = "calcDP(Temp, RH) > DewPoint > TempCalc") + 
  theme_minimal()
```


Visualisation of the temperature calculation against the original temperature and relative humidity. The plot aims to validate the accuracy and consistency of temperature calculations based on relative humidity and dew point.

```{r}
TRHgrid |>
  mutate(
    DewPoint = calcDP(Temp, RH),
    TempCalc = calcTemp(RH, DewPoint), 
    Temp_error = TempCalc - Temp
    ) |>
  ggplot(aes(Temp, RH, z = Temp_error)) + 
  geom_contour_filled(bins = 20) + 
  labs(title = "Difference between calculated temperature vs actual", 
       subtitle = "calcDP(Temp, RH) > DewPoint > TempCalc - Temp (original)") + 
  theme_minimal()

```


### Relative Humidity 

This section investigates relative humidity calculations using derivations from dew point and absolute humidity. Contour plots demonstrate the relationships between calculated and original relative humidity values and the errors. 

```{r}

TRHgrid |>
  mutate(
    DewPoint = calcDP(Temp, RH),
    RH_DPcalc = calcRH_DP(Temp, DewPoint)
    ) |>
  ggplot(aes(Temp, RH, z = RH_DPcalc)) + 
  geom_contour_filled(bins = 20, breaks = seq(0, 100, 5)) + 
  labs(title = "Calculated humidity via dew point vs actual", 
       subtitle = "calcDP(Temp, RH) > DewPoint > RHcalc") + 
  theme_minimal()


TRHgrid |>
  mutate(
    DewPoint = calcDP(Temp, RH),
    RH_DPcalc = calcRH_DP(Temp, DewPoint), 
    RH_DP_error = RH_DPcalc - RH
    ) |>
  ggplot(aes(Temp, RH, z = RH_DP_error)) + 
  labs(title = "Calculated humidity vs actual", 
       subtitle = "calcDP(Temp, RH) > DewPoint > RHcalc - RH (original)") +
  geom_contour_filled(bins = 20) + 
  theme_minimal()


```


```{r}

TRHgrid |>
  mutate(RH_AHcalc = calcRH(Temp, calcAH(Temp, RH))) |>
  ggplot(aes(Temp, RH, z = RH_AHcalc)) + 
  geom_contour_filled(bins = 20, breaks = seq(0, 100, 5)) + 
  labs(title = "Calculated humidity via absolute humidity vs actual", 
       subtitle = "calcAH(Temp, RH) > Absolute humidity > RHcalc - RH (original)") +
  theme_minimal()

TRHgrid |>
  mutate(RH_AH_error = calcRH(Temp, calcAH(Temp, RH)) - RH) |>
  ggplot(aes(Temp, RH, z = RH_AH_error)) + 
  labs(title = "Calculated humidity via absolute humidity vs actual", 
       subtitle = "calcAH(Temp, RH) > Absolute humidity > RHcalc - RH (original)") +
  geom_contour_filled(bins = 20) + 
  theme_minimal()
  
```

Using absolute humidity generates errors at low temperatures and high relative humidities that may become significant or exponentiate if used in machine learning applications. It is not recommended to used the `calcAH` function at low temperature and high relative humidity. 


### Absolute humidity 

```{r}

TRHgrid |>
  mutate(AH = calcAH(Temp, RH)) |>
  ggplot(aes(Temp, RH, z = AH)) + 
  geom_contour_filled(bins = 20) + 
  theme_minimal()

```

### Dew Point

The dew point analysis shows how the dew point temperature varies with changing temperature and relative humidity.

```{r}

TRHgrid |>
  mutate(DP = calcDP(Temp, RH)) |>
  ggplot(aes(Temp, RH, z = DP)) + 
  geom_contour_filled(bins = 20) + 
  theme_minimal()

```


### Water vapour saturation pressure

-   Saturation vapor pressure calculation
-   Actual water vapor pressure determination

```{r}

TRHgrid |>
  mutate(Pws = calcPws(Temp)) |>
  ggplot(aes(Temp, RH, z = Pws)) + 
  geom_contour_filled(bins = 20) + 
  theme_minimal()

```


### Water vapour pressure 



```{r}

TRHgrid |>
  mutate(Pw = calcPw(Temp, RH)) |>
  ggplot(aes(Temp, RH, z = Pw)) + 
  geom_contour_filled(bins = 20) + 
  theme_minimal()

```


### Mixing ratio 

```{r}

TRHgrid |>
  mutate(MR = calcMR(Temp, RH)) |>
  ggplot(aes(Temp, RH, z = MR)) + 
  geom_contour_filled(bins = 20) + 
  theme_minimal()

TRHgrid |>
  mutate(MR_log10 = calcMR(Temp, RH) |> log10()) |>
  ggplot(aes(Temp, RH, z = MR_log10)) + 
  geom_contour_filled(bins = 20) + 
  theme_minimal()

```



### Enthalpy

```{r}

TRHgrid |>
  mutate(Enthalpy = calcEnthalpy(Temp, RH)) |>
  ggplot(aes(Temp, RH, z = Enthalpy)) + 
  geom_contour_filled(bins = 20) + 
  theme_minimal()

TRHgrid |>
  mutate(Enthalpy_log10 = calcEnthalpy(Temp, RH) |> log10()) |>
  ggplot(aes(Temp, RH, z = Enthalpy_log10)) + 
  geom_contour_filled(bins = 20) + 
  theme_minimal()

```


### Preservation Index

```{r}

TRHgrid |>
  mutate(PI = calcPI(Temp, RH)) |>
  ggplot(aes(Temp, RH, z = PI)) + 
  geom_contour_filled(bins = 20) + 
  theme_minimal()

TRHgrid |>
  mutate(PI_log10 = calcPI(Temp, RH) |> log10()) |>
  ggplot(aes(Temp, RH, z = PI_log10)) + 
  geom_contour_filled(bins = 20) + 
  theme_minimal()

```


### Lifetime multiplier

```{r}

TRHgrid |>
  mutate(LifeTime = calcLM(Temp, RH)) |>
  ggplot(aes(Temp, RH, z = LifeTime)) + 
  geom_contour_filled(bins = 20) + 
  theme_minimal()

TRHgrid |>
  mutate(LM_log = calcLM(Temp, RH) |> log10()) |>
  ggplot(aes(Temp, RH, z = LM_log)) + 
  geom_contour_filled(bins = 20) + 
  theme_minimal()

```
