---
title: "Sustainability"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
knitr::opts_chunk$set(echo = FALSE)
library(ConSciR)
library(dplyr)
library(ggplot2)
```

### mydata: Built-in Example Dataset

This dataset contains environmental monitoring data collected from example heritage sites.

It includes measurements of temperature (째C) and relative humidity (%) recorded by sensors over time.

| **Variable** | **Description**                               |
|:-------------|:----------------------------------------------|
| Site         | Location name, e.g., "Museum", "London"       |
| Sensor       | Identifier for the specific sensor            |
| Date         | Timestamp of the measurement (POSIXct format) |
| Temp         | Air temperature in degrees Celsius            |
| RH           | Relative humidity as percentage (0-100%)      |


Example usage to load the dataset from the package:

```{r}
# filepath <- system.file("extdata", "mydata.rda", package = "ConSciR")
# mydata <- load(filepath)

head(mydata)
```

### TRHgrid: Dataset used to visualise the functions

This dataset consists of a grid of temperature and relative humidity values used to visualize function behavior.
Temperatures range from 0째C to 100째C in 0.25째C increments, and relative humidity (RH) ranges from 0% to 100% in 1% increments.
The complete factorial combination of temperature and RH values is created using expand.grid(), where each pair represents a unique condition.

Contour plots are generated to show interactions between derived variables across the temperature-RH grid.
The dataset also supports validation checks for temperature and humidity calculations to understand relationships among the functions.

Example R code to generate the dataset:

```{r}
Temp <- seq(0, 100, 0.25)
RH <- seq(0, 100, 1)

TRHgrid <- 
  expand.grid(Temp, RH) |> 
  tibble() |> 
  rename(Temp = Var1, RH = Var2) 

summary(TRHgrid)
```

```{r}
mydata |> 
  ggplot() +
  geom_line(aes(Date, Temp), col = "red") +
  geom_line(aes(Date, RH), col = "blue") +
  labs(title = "Climate",
       y = "Temperature (red) & Humidity (blue)") + 
  theme_bw()

```

